DOpElib: Differential Equations and Optimization Environment; A Goal Oriented Software Library for Solving PDEs and Optimization Problems

### 2012-11-26 Paper_May_2012/main.tex
Meine Beitraege wie besprochen.

\institute{Christian Goll \at
              Institute of Applied Mathematics, University of Heidelberg
              \\Im Neuenheimer Feld 293/294, 69120 Heidelberg, Germany\\
              Tel.: +49 622-154-6154\\
              \email{cgoll@iwr.uni-heidelberg.de}           %  \\
           \and
Thomas Wick \at
              The Institute for Computational Engineering and Sciences,\\
The University of Texas at Austin,\\
Austin, Texas 78712, USA \\
              Tel.: +1 512-672-7763\\
              \email{twick@ices.utexas.edu}   
 \and
F. Author \at
              first address \\
              Tel.: +123-45-678910\\
              Fax: +123-45-678910\\
              \email{fauthor@example.com}   
}

@@

In many cases, the user is confronted with 
certain bottlenecks. On the one hand, 
in open-source projects for academic research, 
standard high-level routines 
which are necessary to solve even simplest PDEs,
such as linear and nonlinear solvers or time-stepping schemes must be implemented 
by the users themselves. However, such routines are, at large,
problem-independent and could be provided as standard tools. 
As our guiding design principle, we note that in many 
numerical studies, the main task is the computation of some  
target functional, e.g., error norms,
point values, or certain mean-values as drag or lift coefficients.
As a second motivating principle, 
there is nowadays an increasing interest in computing 
optimization problems in which the state equation is governed 
by a PDE. Such optimization problems comprise optimal control,
parameter estimation, inverse problems or optimal experimental design. 
To do this in complex applications usually requires tremendous work already
on the PDE itself before considering optimization of the processes modeled 
by the PDE. Thus it is advantageous to have an environment that allows to 
focus on the development of an appropriate code for the PDE at first but 
asserts due to the required interface, that the code written for the PDE
is then reusable for the application of state-of-the-art optimization 
routines.
So far, there exist a very limited number of open source packages
which fulfill the requirements listed above; 
sucessful in-house code examples are Gascoigne \cite{gascoigne}
and RoDoBo \cite{rodobo}. 
Like DOpElib, both have been initiated as well in 
the Numerical Analysis Group in Heidelberg.

@@

Finally, we note that to compute the requested functional value 
all that is left is to evaluate the functional in the computed solution 
which can be done again using an appropriate integrator method.
Since the sequence of events sketched above is again independent of the 
exact problem we introduce another container around it which we call 
\texttt{StatPDEProblem} that then simply provides the user with the 
method
\begin{lstlisting}
  template<class NEWTON>
     void StatPDEProblem<NEWTON>
            ::ComputeReducedFunctionals();
\end{lstlisting}
The template argument \texttt{NEWTON} represents the choosen solver of the nonlinear problems.
That combines all the necessary steps to calculate $J(u_h)$.

@@

We split the program into the parts \texttt{main.cc}, \texttt{functionals.h} and \texttt{localpde.h}. The \texttt{main.cc}-file is responsible for the declaration of various objects which have mostly to do with the \textit{discretization}, the \textit{solvers} and the general \textit{program sequence}, whereas the other two contain most of the problem-data -- especially the cell-wise contributions of the weak form, the right hand side and the functional. We give a brief explanation on every file in the following.

@@

Now we have everything at hand to create an instantiation of type \texttt{PDEProblemContainer}. This basically represents the discretized problem, and bundles things like boundary conditions, the PDE, the functional and the discretization. First, we give the constructor the \texttt{local\_pde} and the \texttt{dof\_handler}.
\begin{lstlisting}
  PDEProblemContainer<LocalPDE> 
                      prob_container(local_pde,
                                     dof_handler);                                            
\end{lstlisting}
After that, we add the functional under consideration and, because it is a boundary functional, we specify the part of the boundary on which the functional operates (this is done via a 'boundary color', $1$ represents $\Gamma_{circ}$).
\begin{lstlisting}
  prob_container.AddFunctional(&local_functional);
  prob_container.SetBoundaryFunctionalColors(1);
\end{lstlisting}
Next, we incorporate the Dirichlet data, which is given by the class
\texttt{DirichleData}. Then we have to tell the \texttt{prob\_container} for
which components and boundaries we prescribe the Dirichlet conditions. This is
done via a boundary-color in a so-called component mask 
(here, $0$ denotes $\Gamma_D$).
\begin{lstlisting}
  DirichletData dd;
  prob_container.SetDirichletBoundaryColors(0,
                                            comp_mask,
                                            dd);
\end{lstlisting}
Now we create the object which steers the whole solution process, the previously mentioned \texttt{StatPDEProblem}. It takes care of solving the discretized equation and evaluating the given functionals. The class depends actually on some template-parameters, so for the sake of abbreviation, we defined a typedef at the beginning of the \texttt{main.cc}-file.
\begin{lstlisting}

@@

\subsubsection{\texttt{localpde.h}}
We shall now give an overview to the implementation of the specific form of the PDE.

Let $N$ be the number of degrees of freedom of our discretization and $\Set{\phi_{h}^{i}|1\leq i \leq N}$ be a given FE-Basis. As stated in subsection~\ref{subsubsec:stationary problems}, to solve the discrete equation \eqref{eq:discrete_equation} we need the residual, i.e. the vector
\begin{align}\label{eq:discrete_residual}
\left(a(u_h)(\phi_{h}^{i})-(f,\phi_{h}^{i})\right)_{i=1}^N.
\end{align}
The corresponding Jacobian can be defined analogously.
Now all we have to supply here are the cell-wise contributions to these terms, the \texttt{Integrator} takes care of assembling \eqref{eq:discrete_residual} and \eqref{eq:discrete_matrix}.

@@

We show the basic structure of these methods using the example of \texttt{CellEquation}.

@@

In the next lines, we finally implement our concrete PDE. Here, aim to provide 
an interface which is closely related to the weak formulation which 
has to be derived by the user. Thus,
we loop over the quadrature points and the number of DoFs of the cell. In the innermost loop, we evaluate the weak form tested with $\phi_{h}^i$ (represented by \texttt{phi\_i\_grads\_v}) on the cell with the help of the previously selected quadrature formula. 
 \begin{lstlisting}
    for(unsigned int q_point = 0; 
        q_point < n_q_points; 
        q_point++)
    {
      //initialize phi_i_grads_v etc.
      Tensor<2,2> vgrads;
      vgrads[0]= ugrads[q_point][0]; 
      vgrads[1]= ugrads[q_point][1];
      for (unsigned int i=0;i<dofs_per_cell;i++)
        local_cell_vector(i) += 
           ( _nu * vgrads * phi_i_grads_v 
             + //all the other terms)
             * state_fe_values.JxW(q_point);
    }
\end{lstlisting}
\dope{} takes now care of the distribution of the local cell contributions into  the vector of the residual \eqref{eq:residual_vector}. The methods \texttt{CellRightHandSide} and \texttt{CellMatrix} follow the same structure as 
\texttt{CellEquation}, so we do not go into the details here.

\subsubsection{\texttt{localfunctional.h}}
The implementation of the boundary-functional $J$ follows the outline of the
computation of the cell-residual in \texttt{localpde.h}. 
\todo{CG and WW: Pruefen, ob der folgende Absatz nun gut ist:}
The main difference is that as we only want to evaluate a functional and not a
residual-vector, we just loop over the quadrature points and not over the
degrees of freedom on the face. Another deviation, special to the case at
hand, is that we do not have to specify the method \texttt{CellEquation}, but
\texttt{BoundaryValue} instead because $J$ is just a number to be computed. This is due to the fact that we consider a boundary-functional in this example, and so we have to specify the value of the functional not cell-wise but face-wise (for the faces that lie on $\Gamma_{circ}$).

%However, there are two main differences. First, we do not have to specify the method \texttt{CellEquation}, but \texttt{BoundaryValue} instead. This is due to the fact that we consider a boundary-functional in this example, and so we have to specify the value of the functional not cell-wise but face wise (for the faces that lie on $\Gamma_{circ}$). The second difference is that as we only want to evaluate a functional and not a residual-vector, we just loop over the quadrature points and not over the degrees of freedom on the face.
\begin{lstlisting}
 double
 BoundaryValue(const FaceDataContainer& fdc)
  {
    //same as before, get n_q_points etc.
    fdc.GetFaceValuesState("state", ufacevalues);
    fdc.GetFaceGradsState("state", ufacegrads);
\end{lstlisting}
Notice that we get here the values of the computed solution by giving the string "state".
\begin{lstlisting}
    for(unsigned int q_point = 0; 
        q_point < n_q_points; 
        q_point++)
      erg += -1. * ( _nu * ufacegrads[q_point][0]
                * state_fe_face_values.normal_vector(q_point)
                + //other terms)        
                * state_fe_face_values.JxW(q_point);
    return erg;
  }
\end{lstlisting}

@@

It is well-known in fluid mechanics that the pressure and incompressibility terms should be 
treated fully implicitly when time-discretizing 
the Navier-Stokes equations.
For this reason, the implementation is split up
into several items to account for fully implicit 
terms and remaining terms for which we need 
old time-step and present time-step values. 
Hence, in the local cell equation, the parameters
\texttt{scale} and \texttt{scale\_ico}
are used to distinguish these terms:
\begin{lstlisting}
  void CellEquation(...)
   {
   // Usual loops over DoFs and quadrature points
     {
      local_cell_vector(i) += 
         scale * 
         // terms like (\nabla v^{n+1}, \nabla\phi) etc.
        + scale_ico *
         // terms like (-p^{n+1},\nabla\cdot\phi)
     }
   }
\end{lstlisting}
Analogous procedure is performed in the local 
cell matrix.
\begin{remark}
Note that in the stationary case \dope{} automatically sets $ \texttt{scale\_ico} = \texttt{scale}$, so that in fact the user has to implement the \texttt{CellEquation} only once and can henceforth reuse it in all his stationary an instationary problems. This transfers of course also to other methods like \texttt{CellMatrix}.
\end{remark}

In addition to stationary problems, the implementation
of the time-derivative, i.e. the term $ (v^{n+1},\phi)$, must be considered which is done
in the method \texttt{CellTimeEquation} in a straight-forward way. After the initialization of $\texttt{v}$ and the shape function $\texttt{phi\_i}$ and the usual loops over the quadrature points and  degrees of freedom of the cell we fill the local cell vector by
\begin{lstlisting}
        local_cell_vector(i) +=  scale * (v * phi_i) * 
                    state_fe_values.JxW(q_point);    
\end{lstlisting}
Here, the advantage is that \dope{} orders the terms in the time-stepping
problems
such that only $(v^{n+1},\phi)$ must be implemented.

@@

\subsection{Solving of optimization problems}\label{sec:opt}
\todo{For WW: etwas klarer machen, dass wir instationare Optimierung koennen}
Upgrading the framework for solving optimization problems
is described in this section. A prototypical problem reads:
\begin{align*}
\min\;&J(q,u) \\
  &\text{s.t.}\; a(q,u)(\phi) = 0 \quad \forall \phi\in V,\\
  &a \le q \le b,\\
  &g(q,u) \le 0,  
\end{align*}
where $u$ is a FE-function and $q$ can either be a FE-function or some 
fixed number of parameters, $a$ and $b$ are constraint bounds for the control $q$,
and $g(\cdot)$ is some state constraint.
To keep the presentation brief we will neglect the additional bounds
\[
a \le q \le b,\qquad g(q,u) \le 0
\]
although they can be treated by \dope{} as well.
Then, in order to minimize the functional $J$ one may apply the, so-called, 
reduced approach, in which we eliminate the dependent variable $u$ by the solution 
operator $S$ of the PDE, i.e., 
\begin{align*}
a(q,S(q))(\phi) = 0 \quad \forall \phi\in V. 
\end{align*}
This leads to the problem
\begin{align*}
\min\;&j(q) = J(q,S(q)). 
\end{align*}
Then, one can apply any algorithm to solve the, unconstraint, minimization 
problem. This, will in any case require the evaluation of $j(q)$ which 
we can do, e.g., using \texttt{StatPDEProblem::ComputeReducedFunctional} for 
a stationary PDE. If one would like to use derivative based methods, one 
needs in addition derivatives of $j$ and thus of $J$ as well as of $S$. 
To do so, we remark, that one can calculate the derivative of 
$j$ using the adjoint calculus. Then the directional derivative 
of $j$ in direction $\delta q$ is given by
\[
j'(q)(\delta q) = J'_q(q,S(q))(\delta q) + (S^* J_u'(q,S(q)),\delta q)
\]
where the index $q$ or $u$ denote directional derivatives for the 
respective argument. To calculate the action of the adjoint operator $S^*$ 
we note that $z =  S^* J_u'(q,S(q))$
can be computed via the adjoint PDE
\begin{equation}
  \label{opt_KKT_system}
  \begin{aligned}
    a_{u}'(q,u)(\phi , z) &= J'_{u}(q,u)(\phi) & \forall \phi&\in {\cal X}.
  \end{aligned}
\end{equation}
Similarly formulas for the second derivatives can be obtained, see, e.g., 
\cite{BeMeVe06}.

@@

\subsection{Goal-oriented mesh refinement for Navier-Stokes}
In this example we consider a laminar flow around a cylinder in 2d. To be more precise, we are computing the benchmark 2D-1 from \cite{TuSchae96}. The computational domain is as depicted in figure~\ref{fig:example_ns} and the equation is discretized with the Taylor-Hood  element. We want to compute the drag force on the enclosed cylinder
\begin{equation}
J(u) = \frac 1 {20} \int_{\Gamma_{circ}} \nu\partial_nv _1 - pn,
\end{equation}
with the solution $u = (v,p)$, the normal $n$ and the viscosity $\nu = 0.001$.
\begin{figure}[hbt]
\centering
{\includegraphics[width=0.9\textwidth]{Pictures/local_grid_NS.png}}
\caption{Locally adapted grid after eight refinement steps.}
\label{fig:NS_local_grid}
\end{figure}
To achieve an efficient computation of the quantity $J$, we employ goal-oriented mesh refinement with the help of the DWR-method, see \cite{BeRa96}. The method delivers error indicators and by this way computational grids tailored to minimize the error in the previously given functional $J$. The price we have to pay ist the additional computation of a (linear) dual problem, which is implemented as described in section~\ref{sec:opt}. In figure~\ref{fig:NS_local_grid} we see a resulting triangulation after eight refinement steps with roughly $320\,000$ degrees of freedom.

In figure~\ref{fig:NS_comparison} we compare the errors in the functional obtained by using global and local mesh refinement. We see clearly that the local mesh refinement allows for substantial savings with respect to degrees of freedom.

@@

The basic configuration is 
sketched in Figure \ref{fig:example_ns} at which an elastic beam is attached 
behind the rigid cylinder. 
The elastic beam has length
$l=0.35m$ and height $h=0.02m$. The right lower end is positioned at 
$(0.6m,0.19m)$, and
the left end is attached to the circle. 
Control points $A(t)$ (with $A(0) = (0.6,0.2)$) are fixed at the 
trailing edge of the structure, measuring $x$- and $y$-deflections of the beam.

@@

Since the problem is fully nonstationary, the 
dynamics are presents in Figure \ref{res:fsi_2_mesh_and_x_velo}. 
By comparison of our findings in Figure \ref{res:results_ux_and_uy_fsi_2}
with the literature, it can be identified that 
the implementation in \dope{} is correct.

@@

\newpage
\subsection{Compliance minimization}
\todo{For Winni: Achsenbeschriftung, MBB ausschreiben, 
Konfiguration citieren}
In this example, we consider the compliance minimization of a standard MBB-Beam. 
The thickness variable is allowed to take intermediate value apart from $0$ and $1$,
i.e., we consider a variable thickness sheet. The discretization is done 
using $Q_2$ elements for the displacement and discontinuous $P_0$ elements for the 
thickness.
\begin{figure}
\centering
\includegraphics[width=1.\textwidth, viewport=150 0 1024 500, clip]{Pictures/MBB.png}
\caption{Thickness distribution for the MBB-beam minimum compliance problem.} 
\label{res:mbb}
\end{figure}
A typical thickness distribution is shown in Figure~\ref{res:mbb}.

\section{Conclusions}
\todo{Frage an CG und Winni: Guter Abschluss?}\todocg
In this article, we described the architecture 
of the novel software library DOpElib for solving 
PDEs and optimization problems.  The key idea behind 
the algorithms is the computation of some target 
functional \todocg{Ich wuerde eventuell die userfreundlichkeit mehr in den Fokus ruecken? The key idea of the software is to supply a toolkit which allows for fast implementation of the problems at hand, but at the same time gives the user full flexibility to change or alter the used algorithms}. This is achieved by a unified interface 
to state of the art time-stepping methods, nonlinear solvers and optimization
routines. The capabilites of DOpElib are demonstrated with the help of three examples.

>>>>

\institute{Christian Goll \at
              Institute of Applied Mathematics, University of Heidelberg
              \\Im Neuenheimer Feld 293/294, 69120 Heidelberg, Germany\\
              Tel.: +49 622-154-6154\\
              \email{cgoll@iwr.uni-heidelberg.de}           %  \\
           \and
Thomas Wick \at
              The Institute for Computational Engineering and Sciences,\\
The University of Texas at Austin,\\
Austin, Texas 78712, USA \\
              Tel.: +1 512-672-7763\\
              \email{twick@ices.utexas.edu}   
 \and
W. Wollner \at
              Universit\"at Hamburg\\
              Department of Mathematics\\
              \email{winnifried.wollner@math.uni-hamburg.de}   
}

@@

In many cases, the user is confronted with 
certain bottlenecks. On the one hand, 
in open-source projects for academic research, 
standard high-level routines 
which are necessary to solve even simplest PDEs,
such as linear and nonlinear solvers or time-stepping schemes must be implemented 
by the users themselves. However, such routines are, at large,
problem-independent and could be provided as standard tools. 
As our guiding design principle, we note that in many 
numerical studies, the main task is the computation of some  
target functional, e.g., error norms,
point values, or certain mean-values as drag or lift coefficients.
As a second motivating principle, 
there is nowadays an increasing interest in computing 
optimization problems in which the state equation is governed 
by a PDE. Such optimization problems comprise optimal control,
parameter estimation, inverse problems or optimal experimental design. 
To do this in complex applications usually requires tremendous work already
on the PDE itself before considering optimization of the processes modeled 
by the PDE. Thus it is advantageous to have an environment that allows to 
focus on the development of an appropriate code for the PDE at first but 
asserts due to the required interface, that the code written for the PDE
is then reusable for the application of state-of-the-art optimization 
routines.
So far, there exist a very limited number of open source packages
which fulfill the requirements listed above; 
successful in-house code examples are Gascoigne \cite{gascoigne}
and RoDoBo \cite{rodobo}. 
Like DOpElib, both have been initiated as well in 
the Numerical Analysis Group in Heidelberg.

@@

Finally, we note that to compute the requested functional value 
all that is left is to evaluate the functional in the computed solution 
which can be done again using an appropriate integrator method.
Since the sequence of events sketched above is again independent of the 
exact problem we introduce another container around it which we call 
\texttt{StatPDEProblem} that then simply provides the user with the 
method
\begin{lstlisting}
  template<class NEWTON>
     void StatPDEProblem<NEWTON>
            ::ComputeReducedFunctionals();
\end{lstlisting}
The template argument \texttt{NEWTON} represents the chosen solver of the nonlinear problems.
That combines all the necessary steps to calculate $J(u_h)$.

@@

We split the program into the parts \texttt{main.cc}, \texttt{func.h} and \texttt{pde.h}. The \texttt{main.cc}-file is responsible for the declaration of various objects which have mostly to do with the \textit{discretization}, the \textit{solvers} and the general \textit{program sequence}, whereas the other two contain most of the problem-data -- especially the cell-wise contributions of the weak form in \texttt{pdh.h} and the functional in \texttt{func.h}. 
We give a brief explanation on every file in the following.

@@

Now we have everything at hand to create an instantiation of the container for the problem data \texttt{PDEProblemContainer}. This basically represents the discretized problem, and bundles things like boundary conditions, the PDE, the functional and the discretization. First, we give the constructor the \texttt{local\_pde} and the \texttt{dof\_handler}.
\begin{lstlisting}
  PDEProblemContainer<LocalPDE> 
                      prob_container(local_pde,
                                     dof_handler);                                            
\end{lstlisting}
After that, we add the functional under consideration and, because it is a boundary functional, we specify the part of the boundary on which the functional operates (this is done via a 'boundary color', $1$ represents $\Gamma_{circ}$).
\begin{lstlisting}
  prob_container.AddFunctional(&local_functional);
  prob_container.SetBoundaryFunctionalColors(1);
\end{lstlisting}
Next, we incorporate the Dirichlet data. To this end, we instantiate a 
class \texttt{DirichletData} containing the required description of the 
boundary values. Then we have to tell the \texttt{prob\_container} for
which components and boundaries we prescribe the Dirichlet conditions. This is
done via a boundary-color in a so-called component mask 
(here, $0$ denotes $\Gamma_D$).
\begin{lstlisting}
  DirichletData dd;
  prob_container.SetDirichletBoundaryColors(0,
                                            comp_mask,
                                            dd);
\end{lstlisting}
Now we create the object which steers the whole solution process, the previously mentioned \texttt{StatPDEProblem}. It takes care of solving the discretized equation and evaluating the given functionals. The class depends actually on some template-parameters, so for the sake of abbreviation, we defined a typedef at the beginning of the \texttt{main.cc}-file.
\begin{lstlisting}

@@

\subsubsection{\texttt{pde.h}}
We shall now give an overview to the implementation of the specific form of the PDE.

Let $N$ be the number of degrees of freedom of our discretization and $\Set{\phi_{h}^{i}|1\leq i \leq N}$ be a given FE-Basis. As stated in subsection~\ref{subsubsec:stationary problems}, to solve the discrete equation \eqref{eq:discrete_equation} we need the residual, i.e. the vector
\begin{align}\label{eq:discrete_residual}
\left(a(u_h)(\phi_{h}^{i})-(f,\phi_{h}^{i})\right)_{i=1}^N.
\end{align}
The corresponding Jacobian can be defined analogously.
Now all we have to supply here are the cell-wise contributions to these terms, the \texttt{Integrator} takes care of assembling \eqref{eq:discrete_residual}
and similarly of the matrix for the linearized equation to be solved.

@@

We show the basic structure of these methods using the \texttt{CellEquation} 
method for the implementation of the element-wise integrals.

@@

In the next lines, we finally implement our concrete PDE. Here, aim to provide 
an interface which is closely related to the weak formulation which 
has to be derived by the user. Thus,
we loop over the quadrature points and the number of DoFs of the cell. In the innermost loop, we evaluate the weak form tested with $\phi_{h}^i$ (represented by \texttt{phi\_i\_grads\_v}) on the cell with the help of the previously selected quadrature formula. 
 \begin{lstlisting}
    for(unsigned int q_point = 0; 
        q_point < n_q_points; 
        q_point++)
    {
      //initialize phi_i_grads_v etc.
      Tensor<2,2> vgrads;
      vgrads[0]= ugrads[q_point][0]; 
      vgrads[1]= ugrads[q_point][1];
      for (unsigned int i=0;i<dofs_per_cell;i++)
        local_cell_vector(i) += 
           ( _nu * vgrads * phi_i_grads_v 
             + //all the other terms
           ) * fe_values.JxW(q_point);
    }
\end{lstlisting}
\dope{} takes now care of the distribution of the local cell contributions into  the vector of the residual \eqref{eq:residual_vector}. The methods \texttt{CellRightHandSide} and \texttt{CellMatrix} follow the same structure as 
\texttt{CellEquation}, so we do not go into the details here.

\subsubsection{\texttt{localfunc.h}}
The implementation of the boundary-functional $J$ follows the outline of the
computation of the cell-residual in \texttt{localpde.h}. 
\todo{Habe ich mal umgeschrieben, bitte nochmal ansehen~WW.}
The main difference, and common to all functionals is that as we only 
want to evaluate a functional, i.e., calculate a single number, and not a
residual-vector, i.e., calculate the elements of a vector. Thus, we 
just loop over the quadrature points and not over the degrees of freedom 
on the face. In the present situation, we want to evaluate a functional 
that consists of a boundary integral. Thus we need to evaluate 
integrals on faces, that lie on $\Gamma_{circ}$, instead of integrals over
elements. 
This face contributions are calculated in \texttt{BoundaryValue} and added to 
sum up to the value of $J$.

\begin{lstlisting}
 double
 BoundaryValue(const FaceDataContainer& fdc)
  {
    //same as before, get n_q_points etc.
    const auto& face_values = cdc.GetFEFaceValuesState();
    fdc.GetFaceValuesState("state", ufacevalues);
    fdc.GetFaceGradsState("state", ufacegrads);
\end{lstlisting}
Notice that we get here the values of the computed solution by giving the string "state".
\begin{lstlisting}
    for(unsigned int q_point = 0; 
        q_point < n_q_points; 
        q_point++)
      erg -= ( _nu * ufacegrads[q_point][0]
               * face_values.normal_vector(q_point)
               + //other terms        
             ) * face_values.JxW(q_point);
    return erg;
  }
\end{lstlisting}

@@

It is well-known in fluid mechanics that the pressure and incompressibility terms should be 
treated fully implicitly when time-discretizing 
the Navier-Stokes equations.
For this reason, the implementation is split up
into several items to account for fully implicit 
terms and remaining terms for which we need 
old time-step and present time-step values. 
Hence, in the local cell equation, the parameters
\texttt{scale} and \texttt{scale\_ico}
are used to distinguish these terms:
\begin{lstlisting}
  void CellEquation(...)
   {
   // Usual loops over DoFs and quadrature points
     {
      local_cell_vector(i) += 
         scale * 
         // terms like (\nabla v^{n+1}, \nabla\phi) etc.
        + scale_ico *
         // terms like (-p^{n+1},\nabla\cdot\phi)
     }
   }
\end{lstlisting}
Analogous procedure is performed in the local 
cell matrix.
\begin{remark}
Note that in the stationary case there is no such difference and accordingly 
\dope{} automatically sets $ \texttt{scale\_ico} = \texttt{scale}$, so that in fact the user has to implement the \texttt{CellEquation} only once and can henceforth reuse it in all his stationary and nonstationary problems. This transfers of course also to other methods like \texttt{CellMatrix}.
\end{remark}

In comparison to stationary problems, we need to consider, in addition, 
the implementation
of the time-derivative, i.e., the term $ (v^{n+1},\phi)$. This is done
in the method \texttt{CellTimeEquation} in a straight-forward way. After the initialization of $\texttt{v}$ and the shape function $\texttt{phi\_i}$ and the usual loops over the quadrature points and  degrees of freedom of the cell we fill the local cell vector by
\begin{lstlisting}
    local_cell_vector(i) +=  scale * (v * phi_i)  
                             * fe_values.JxW(q_point);    
\end{lstlisting}
Here, the advantage is that \dope{} orders the terms in the time-stepping
problems
such that only $(v^{n+1},\phi)$ must be implemented.

@@

\subsection{Solving of optimization problems}\label{sec:opt}
Once we are able to solve stationary and nonstationary PDEs
we can consider optimization problems constrained by these PDEs.
How this is done will be described in this section.

A prototypical PDE constrained optimization problem reads:
\begin{align*}
\min\;&J(q,u) \\
  &\text{s.t.}\; a(q,u)(\phi) = 0 \quad \forall \phi\in V,\\
  &q_- \le q \le q_+,\\
  &g(q,u) \le 0,  
\end{align*}
where $u$ is a function and $q$ can either be a function or some 
fixed number of parameters. the numbers $q_-$ and $q_+$ are bounds for 
the control $q$, and $g(\cdot)$ is some state constraint.
To keep the presentation brief we will neglect the additional bounds
\[
a \le q \le b,\qquad g(q,u) \le 0
\]
although they can be treated by \dope{} as well.
Then, in order to minimize the functional $J$ one may apply the, so-called, 
reduced approach, in which we eliminate the dependent variable $u$ by the solution 
operator $S$ of the (stationary or nonstationary) PDE, i.e., 
\begin{align*}
a(q,S(q))(\phi) = 0 \quad \forall \phi\in V. 
\end{align*}
This leads to the problem
\begin{align*}
\min\;&j(q) = J(q,S(q)). 
\end{align*}
This can be discretized using the finite element approximation $S_h$ of $S$ 
and yields the problem 
\begin{align*}
\min\;&j_h(q) = J(q,S_h(q)). 
\end{align*}
After possibly discretizing the control $q$ as well using finite elements, 
one can apply any algorithm to solve the, unconstraint, minimization 
problem. This, will in any case require the evaluation of $j_h(q)$ which 
we can do, e.g., using \texttt{StatPDEProblem::ComputeReducedFunctional} for 
a stationary PDE or using \texttt{InstatPDEProblem::ComputeReducedFunctional} 
for nonstationary PDEs.

If one would like to use derivative based methods, one 
needs in addition derivatives of $j_h$ and thus of $J$ as well as of $S_h$. 
To do so, we remark, that one can calculate the derivative of 
$j$ using the adjoint calculus. Then the directional derivative 
of $j$ in direction $\delta q$ is given by
\[
j'_h(q)(\delta q) = J'_q(q,S_h(q))(\delta q) + (S_h^* J_u'(q,S_h(q)),\delta q)
\]
where the index $q$ or $u$ denote directional derivatives for the 
respective argument. To calculate the action of the adjoint operator $S_h^*$ 
we note that $z =  S_h^* J_u'(q,S_h(q))$
can be computed via the adjoint PDE
\begin{equation}
  \label{opt_KKT_system}
  \begin{aligned}
    a_{u}'(q,u)(\phi , z) &= J'_{u}(q,u)(\phi) & \forall \phi&\in {\cal X}.
  \end{aligned}
\end{equation}
Similarly formulas for the second derivatives can be obtained, see, e.g., 
\cite{BeMeVe06}.

@@

\subsection{Goal-oriented mesh refinement for Navier-Stokes}
In this example we consider a laminar flow around a cylinder in 2d. To be more precise, we are computing the benchmark 2D-1 from \cite{TuSchae96}. The computational domain is as depicted in Fig.~\ref{fig:example_ns} and the equation is discretized with the Taylor-Hood  element. We want to compute the drag force on the enclosed cylinder
\begin{equation}
J(u) = \frac 1 {20} \int_{\Gamma_{circ}} \nu\partial_nv _1 - pn,
\end{equation}
with the solution $u = (v,p)$, the normal $n$ and the viscosity $\nu = 0.001$.
\begin{figure}[hbt]
\centering
{\includegraphics[width=0.9\textwidth]{Pictures/local_grid_NS.png}}
\caption{Locally adapted grid after eight refinement steps.}
\label{fig:NS_local_grid}
\end{figure}
To achieve an efficient computation of the quantity $J$, we employ goal-oriented mesh refinement with the help of the DWR-method, see \cite{BeRa96}. The method delivers error indicators and by this way computational grids tailored to minimize the error in the previously given functional $J$. The price we have to pay is the additional computation of a (linear) dual problem, which is implemented as described in section~\ref{sec:opt}. In Fig.~\ref{fig:NS_local_grid} we see a resulting triangulation after eight refinement steps with roughly $320\,000$ degrees of freedom.

In Fig.~\ref{fig:NS_comparison} we compare the errors in the functional obtained by using global and local mesh refinement. We see clearly that the local mesh refinement allows for substantial savings with respect to degrees of freedom.

@@

The basic configuration is 
sketched in Fig.~\ref{fig:example_ns} at which an elastic beam is attached 
behind the rigid cylinder. 
The elastic beam has length
$l=0.35m$ and height $h=0.02m$. The right lower end is positioned at 
$(0.6m,0.19m)$, and
the left end is attached to the circle. 
Control points $A(t)$ (with $A(0) = (0.6,0.2)$) are fixed at the 
trailing edge of the structure, measuring $x$- and $y$-deflections of the beam.

@@

Since the problem is fully nonstationary, the 
dynamics are presents in Fig.~\ref{res:fsi_2_mesh_and_x_velo}. 
By comparison of our findings in Fig.~\ref{res:results_ux_and_uy_fsi_2}
with the literature, it can be identified that 
the implementation in \dope{} is correct.

@@

\newpage
\subsection{Compliance minimization}
In this example, we consider the compliance minimization of a standard MBB-Beam (Messerschmidt-B\"olkow-Blohm-Beam), see, e.g.,~\cite{BendsoeSigmund:2003}.
The thickness variable is allowed to take intermediate value apart from $0$ and $1$,
i.e., we consider a variable thickness sheet. The discretization is done 
using $Q_2$ elements for the displacement and discontinuous $P_0$ elements for the 
thickness.
\begin{figure}
\centering
\includegraphics[width=1.\textwidth, viewport=205 88 975 500, clip]{Pictures/MBB.png}
\caption{Thickness distribution for the MBB-beam minimum compliance problem.} 
\label{res:mbb}
\end{figure}
A typical thickness distribution is shown in Fig.~\ref{res:mbb}.

\section{Conclusions}
\todo{Frage an CG und Winni: Guter Abschluss?}
In this article, we described the architecture 
of the novel software library DOpElib for solving 
PDEs and optimization problems.  The key idea behind 
the algorithms is the computation of some target 
functional \todocg{Ich wuerde eventuell die userfreundlichkeit mehr in den Fokus ruecken? The key idea of the software is to supply a toolkit which allows for fast implementation of the problems at hand, but at the same time gives the user full flexibility to change or alter the used algorithms}\todo{Bin daf\"ur~WW}. This is achieved by a unified interface 
to state of the art time-stepping methods, nonlinear solvers and optimization
routines. The capabilities of DOpElib are demonstrated with the help of three examples.
